Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-05-13 manuscript.Rmd
add gdl's corrections

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot*, Marguerite Cogné*, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
date: "2024-03-19"
output: 
  officedown::rdocx_document:
    base_format: "bookdown::word_document2"
    reference_docx: word-styles-ref-02.docx
bibliography: references.bib
nocite: |
  @Orsini1976, @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
---

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural land and diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFSs) in Timor-Leste, namely: cropping systems with fallow, silvopastures, young agroforests, home gardens and forest gardens. Our results show that these traditional AFSs can store large amounts of carbon: they stored an average of 156\ Mg\ C\ ha$^{-1}$, and the AFSs with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, close to the values of old-growth tropical forests. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFSs in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste's climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are both more resilient to climate change, while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the amount of carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of the food production system [@Lal2004;@Minasny2017].

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation [@Albrecht2003;@Verchot2007;@Lasco2014;@Cardinael2021]. Agroforestry is a collective name for land use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals, in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Agroforestry allows for diversification of agricultural production, making agricultural systems more resilient [@Duffy2021;@Terasaki2023].

Agroforestry systems (AFSs) have the potential to store large amounts of carbon in the vegetation and soil, and this carbon storage capacity depends on the characteristics of AFS considered [@Albrecht2003;@Feliciano2018]. As trees can store large amounts of carbon, the carbon stored in the vegetation of AFSs is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFSs on soil organic carbon (SOC) has been relatively little studied [@Beillouin2021], is more variable and depends on a number of factors. Although AFSs appear to increase SOC stocks overall, some AFS such as woodlots (i.e. planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFSs, while AFSs such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type can also influence SOC in AFSs, which may explain the variability of observations reported in the literature [@Feliciano2018;@Ma2020;@Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS could only be evidenced for the topsoil layer [< 10 cm; @Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods, and food security is threatened by population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime [@WB2009;@Molyneux2012]. In this context, productive, resilient and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008;@Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFSs. A precise quantification of these carbon stocks would improve the understanding of the ecological importance of these AFSs at both local and global scales, and help to integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFSs in Timor-Leste by quantifying the carbon storage in both biomass and soil of five traditional AFSs: cropping systems with fallow, silvopastures, young agroforests, home gardens and forest gardens. We hypothesize that different types of traditional AFSs will exhibit measurable differences in aboveground carbon (AGC; H1a) and SOC (H1b) stocks. We anticipate, following @Ma2020, that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; H2a), but no relationship with SOC below 10 cm (H2b).

@@

site_summary = site_env[, .(
  coord = paste(round(mean(long),2), round(mean(lat),2), sep = ", "), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))), collapse = " - ")),
  .(site)]
```

The research was carried out in the center of the Baucau municipality, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m above sea level (asl); and (ii) the village of Osso-Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso-Luga", coord]`) located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso-Luga", elev]` m asl (Figure\ \@ref(fig:map-sites)). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differed between villages (Table\ S\@ref(tab:site-soil)): Gariuai had a clay loam texture whereas Osso-Luga had a clay texture. The two villages also have different recent histories: Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. During the period of abandonment, the AFSs in Osso-Luga lay fallow.

@@

We used the typology described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair et al. (1993, 2021) and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFSs identified were: Crop system including a wooded Fallow phase (CF), Silvopasture (SP), Young Agroforest (YA), Home Garden (HG) and Forest Garden (FG). CFs are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SPs consist in a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YAs are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years. HGs are plots of land located close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFSs are presented in more detail as supplemental information (Table\ S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected 3 plots from each of the 5 traditional AFSs identified in each of the two sites (30 plots in total; Figure\ \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFSs except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Figure\ S\@ref(fig:design)): within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were done with a dendrometer.

Taxonomic identification was carried out using Tetum  (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory government's Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

In each of the 30 sample plots, three composite samples were taken from three of the five 10 x 10 m subplots, following a transect covering the intra-plot variability and at a minimum distance of 1 m from any tree (Figure\ S\@ref(fig:design)). Soils were sampled with a 15cm auger at 0-10 cm and 20-30 cm depth; due to the size of the auger we couldn't accurately sample the intermediate soil layer. Soils samples were then air-dried and sieved at 2 mm. They were subjected to standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table\ S\@ref(tab:site-soil)). Among these analyses, total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm$^3$ at 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until reaching a stable dry mass, sieved to 2 mm, and then the fine fraction (< 2 mm) was weighed to be used in SOC stocks calculation (equation \@ref(eq:soc)).

@@

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as a cylinder, using the height and diameter measured *in situ*, with a volumetric shrinkage of 11% [@WD2007]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

Estimated AGC and SOC stocks were log-transformed for all statistical tests, because we expected them to be log-normally distributed as they only take positive values. We removed one CF plot at Gariuai from the following statistical tests, as it did not contain any trees exceeding the minimum measurement size (CBH = 30 cm) and its AGC stocks were therefore estimated at 0\ Mg\ C\ ha$^{-1}$ (even if there were trees < 30 cm CBH) and could not be log-transformed. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo function from the R package BIOMASS, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013]; large and small errors on 5 and 95% of trees respectively [@Chave2004;@Rejou-Mechain2017].
To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (as recommended by Cirad Laboratories in Montpellier, France), 10% for the mass of the fine fraction [@Page1999].

@@

Total carbon stocks ranged between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks ranged between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The proportion of variance explained by the linear mixed models were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths respectively, but a marginal R$^2$ of just `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`% respectively when only the fixed effect was considered (Figure\ \@ref(fig:plot-lmereg)).

@@

## Traditional AFS in Timor Leste can store large C stocks

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot^*^, Marguerite Cogné^*^, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
date: ""
output: 
  officedown::rdocx_document:
    base_format: "bookdown::word_document2"
    reference_docx: word-styles-ref-02.docx
bibliography: references.bib
nocite: |
  @Orsini1976, @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
---

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural land and diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. Our results show that these traditional AFS can store large amounts of carbon, with the average being 156\ Mg\ C\ ha$^{-1}$. The AFS with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, close to the values of old-growth tropical forests. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as the world’s population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are more resilient to climate change while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the amount of carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of food production systems [@Lal2004;@Minasny2017].

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation [@Albrecht2003;@Verchot2007;@Lasco2014;@Cardinael2021]. Agroforestry is a collective name for land use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals, in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Agroforestry allows for the diversification of agricultural production, making agricultural systems more resilient [@Duffy2021;@Terasaki2023].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation and soil, and this carbon storage capacity depends on the characteristics of the AFS considered [@Albrecht2003;@Feliciano2018]. As trees can store large amounts of carbon, the carbon stored in the vegetation of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) has been relatively little studied [@Beillouin2021], is more variable and depends on several factors. Although AFS appear to increase SOC stocks overall, some AFS such as woodlots (i.e., planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFS, while AFS such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type can also influence SOC in AFS, which may explain the variability of observations reported in the literature [@Feliciano2018;@Ma2020;@Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS could only be evidenced for the topsoil layer [< 10 cm; @Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009;@Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008;@Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks would improve understanding of the ecological importance of these AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying the carbon storage in both biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. We hypothesize that there are measurable differences in aboveground carbon (AGC; H1a) and SOC (H1b) stocks between these different types of traditional AFS. In line with Ma et al. (2020), we anticipate that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; H2a), but no relationship with SOC below 10 cm (H2b).

@@

site_summary = site_env[, .(
  coord = paste(round(mean(long),2), round(mean(lat),2), sep = ", "), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))/10)*10, collapse = " and ")),
  .(site)]
```

The research was carried out in the center of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (asl). The second, the village of Osso-Luga in Samalari *suco* (long-lat coordinates: `r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` m asl (Figure\ \@ref(fig:map-sites)). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table\ S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different recent histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment.

@@

We used the typology described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair et al. (1993, 2021) and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFS identified were as follows: Crop systems including a wooded Fallow phase (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). CF are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts, and squash) which are left fallow after one to three years. SP are vast areas (generally 200 to 500 ha) where cattle are grazed under the shade of trees. YA are agricultural plots where farmers plant rows of trees spaced approximately 5 x 10 m apart, which they maintain for two to ten years. HG are plots of land located close to houses where crops and livestock (poultry, pigs, cattle, etc.) are kept in the shade of numerous trees and palms for multiple uses (wood, fruit, fiber, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFS are presented in more detail in the supplemental information section (Table\ S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected three plots from each of the five traditional AFS identified in each of the two sites (30 plots in total; Figure\ \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Figure\ S\@ref(fig:design)). Within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were taken with a dendrometer.

Taxonomic identification was carried out using Tetum  (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory Government's Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

In each of the 30 sample plots, three composite samples were taken from three of the five 10 x 10 m subplots, following a transect covering the intra-plot variability and at a minimum distance of 1 m from any tree (Figure\ S\@ref(fig:design)). Soils were sampled with a 15 cm auger at 0-10 cm and 20-30 cm depth; due to the size of the auger we could not accurately sample the intermediate soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table\ S\@ref(tab:site-soil)). Total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm3 at 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until a stable dry mass was reached and then sieved to 2 mm. The fine fraction (< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

We estimated tree-level AGC stocks from the tree CBH and height measurements using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as a cylinder, using the height and diameter measured *in situ*, with a volumetric shrinkage of 11% [@WD2007]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

The estimated AGC and SOC stocks were log-transformed for all statistical tests because, as they only take positive values, we expected them to be log-normally distributed. We removed one CF plot at Gariuai from the following statistical tests, as it did not contain any trees exceeding the minimum measurement size (CBH = 30 cm) and its AGC stocks were therefore estimated at 0\ Mg\ C\ ha$^{-1}$ (even if there were trees < 30 cm CBH) and could not be log-transformed. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo function from the R package BIOMASS, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013] and large and small errors for 5 and 95% of trees, respectively [@Chave2004;@Rejou-Mechain2017]. To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (as recommended by Cirad laboratories in Montpellier, France), and 10% for the mass of the fine fraction [@Page1999].

@@

Total carbon stocks ranged between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks were the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks were the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks ranged between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)), but not between sites (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The AFS with the lowest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The conditional R^2^ (i.e., the proportion of variance explained by the linear mixed models) were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths, respectively, but the marginal R^2^ (i.e., when considering only the fixed effect) were only `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`%, respectively (Figure\ \@ref(fig:plot-lmereg)).

@@

## Traditional AFS in Timor-Leste can store large C stocks

Compared to values reported in the literature for other AFS, the carbon stocks in this study were relatively high. Specifically, the average SOC stocks (`r dataC_plot[variable=="SOC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$) were higher than other reported values. In a meta-analysis by @Shi2018, it was estimated that AFS stored around 50\ Mg\ C\ ha$^{-1}$ at a 0-30 cm depth. Such large SOC stocks could be due to the high soil pH associated with high exchangeable Ca content at our sites (Table\ \@ref(tab:site-soil)), as Ca is frequently attributed to enhanced organic matter stabilization [@Wiesmeier2019]. In contrast, the average biomass stocks (`r dataC_plot[variable%in%c("AGC", "BGC"), sum(mean), .(plot)][, round(mean(V1))]`\ Mg\ C\ ha$^{-1}$) were similar to the average of 74\ Mg\ C\ ha$^{-1}$ estimated by @Ma2020, although some AFS had particularly high biomass values. The AFS with the highest carbon stocks, `r dataC_afs[variable=="TOT", afs[which.max(mean)]]`, had total carbon stocks (`r dataC_afs[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$) close to those of old-growth tropical forests [270\ Mg\ C\ ha$^{-1}$, @Lal2005]. The high carbon stocks in FG can be attributed to the length of time they had been established, which was generally longer than 50 years, with little human intervention (Table\ S\@ref(tab:afs-typology)). In terms of structure and carbon stocks, FG should therefore resemble old secondary forests, which have been shown to recover more than 60% of the AGC stocks and almost all the SOC stocks of an old forest [@Poorter2021]. In addition, FG contain a large number of trees of the same age that were planted when the AFS was established and can survive to reach significant sizes, similar to forest plantations that can grow to old-growth-forest levels of AGC stocks after only a few decades [@Brown2020;@Njoukam1996].

Although agroforestry has been identified as one of the possible levers to increase carbon storage in the Timorese agricultural landscape [@UNFCCC2020;@Paudel2022], to our knowledge no study has quantified carbon stocks in these systems. Our study could therefore help to better quantify the national contribution of agroforestry to Timor-Leste's climate change mitigation commitments [@UNFCCC2020] and the potential carbon gains from a wider adoption of these techniques. Further studies should help to increase the representativeness of the AFS sampled by including a greater diversity of AFS and environmental conditions. Another critical aspect is the quantification of temporal changes in carbon stocks, which could be assessed in future studies either by monitoring chronosequences of plots where the time since installation is known, or by remeasuring plots over time.
